Using Hybrid Machine Learning for Fundamental Stellar Parameter Inference

### 2024-05-07 paper/paper2_neo/paper.tex
Add bibliography and citations to paper

\documentclass[twocolumn]{aastex631}
\usepackage{graphicx}
\usepackage{physics}
\usepackage[super]{nth}
\usepackage{enumitem}
\usepackage{natbib}

\graphicspath{{./images/}}

@@

\begin{abstract}
    The advent of machine learning methods has revolutionized 
    numerous scientific methods, among which is spectroscopic 
    inference. Studies can and have shown that machine learning 
    methods prove effective in inferring what we call
    fundamental stellar parameters. These include $T_{\mathrm{eff}}$,
    $\log(g)$, and [Fe/H] for the purposes of this study but can 
    also encompass other parameters such as [$\alpha$/Fe], C/O, and
    $f_{sed}$. While current methods tend to treat spectra purely
    from a data perspective, we propose to use a hybrid machine 
    learning (HML) method that uses the semi-empirical approach 
    of treating spectra as a set of spectral lines put forth by 
    \texttt{blase}, combined with traditional linear interpolation 
    as well as Bayesian optimization. We interpolate the four 
    line properties ($\mu'$, $\ln(a)$, $\sigma$, $\gamma$) of 
    128,723 unique spectral lines across a subset of the PHOENIX
    synthetic model grid to create the PHOENIX generator, allowing
    us to reconstruct spectra at any continuously valued point 
    lying within the domain of the PHOENIX subset, and even reconstruct 
    batches of spectra in pseudo-parallel fashion, with the speedup 
    exceeding 20 times once more than 60 spectra are reconstructed. 
    We then combine the PHOENIX generator with Bayesian optimization to 
    efficiently infer fundamental stellar parameters. This study is 
    however a proof of concept, as we have not yet included various
    features that would be necessary for a full-fledged tool for 
    spectroscopic inference, however we show that semi-empirical 
    HML methods are a viable, albeit fledgling alternative to 
    more traditional approaches.
\end{abstract}

@@

Scientists want to extract as much information as possible from
a star's spectra to get back the properties that lie at the root of it all. 
However, this has proven to be no easy task. Spectroscopic surveys such as 
APOGEE develop their own in-house pipelines to extract stellar parameters
from spectra (ASPCAP in the case of APOGEE), and these pipelines are fairly
effective. However, they all work on the core assumption that the data to be 
analyzed is a list of pixels. In addition, the pipelines that surveys 
develop are typically closed-source and usually limited to the scope of the 
survey itself, causing a sort of fragmentation. This motivates the 
development of a more universal, open-source tool that can do what spectral 
analysis pipelines do, but in an alternative way.

Multiple efforts have been made in this direction, treating spectra in 
different ways. The standard practice is to treat the wavelength and flux as 
simply two arrays and implement any from a myriad of different algorithms to 
inference fundamental stellar parameters. Others take a more mathematical 
approach by decomposing spectra into some eigenbasis and treating it as a 
regression problem, such as \texttt{starfish}. However, \texttt{blase} takes 
a different approach. It treats spectra not as a set of pixels or a set of 
eigenbasis coefficients but as a set of spectral lines, specifically Voigt 
profiles. There is no metric of this approach being better or worse than the 
others mentioned so far, but it is a different paradigm and is worth 
exploring as a semi-empirical tool. This study aims to take advantage of 
\texttt{blase}'s ability to change a spectrum's basis from flux versus 
wavelength into spectral line versus spectral line properties, and use that 
to develop a robust tool for stellar parameter inference.

@@

We chose the PHOENIX synthetic spectral model grid as the basis for this 
study, because it is fairly well-known. While we plan to extend this tool to 
leverage information across multiple model grids in the future, both for 
accuracy and to increase the scope to include substellar objects with grids
such as Sonora, for now we limit our scope to a subset of the PHOENIX grid.

@@

\begin{table}[h!]
    \centering
    \begin{tabular}{lll}
        \hline
        \bf{Parameter} & \bf{Symbol} & \bf{Range}\\
        \hline
        Alpha Element Abundance & $\alpha$ & [0] dex\\
        Iron Abundance & [Fe/H] & [-0.5, 0] dex\\
        Effective Temperature & $T_{\mathrm{eff}}$ & [2300, 12000] K\\
        Surface Gravity & $\log(g)$ & [2, 6]\\
        Wavelength & $\lambda$ & [8038, 12849] \AA\\
        \hline
    \end{tabular}
    \caption{The subset of the PHOENIX grid used in this study. These limits 
    were imposed to reduce the computational cost of the algorithms and to 
    ensure a rectilinear parameter space in order to work with \texttt{scipy}'s
    \texttt{RegularGridInterpolator}. The wavelength limits in particular 
    roughly line up with that of the Habitable Zone Planet Finder (HPF) 
    spectrograph. This subset is comprised of 1314 individual spectra.}
\end{table}

\subsection{Preprocessing with \texttt{gollum}}
First, the PHOENIX subset was accessed directly from the PHOENIX website
using \texttt{gollum}'s download option. The spectra were then put through a 
three-step preprocessing pipeline.
\begin{enumerate}
    \item \textit{Blackbody Division}: Since the $T_{\mathrm{eff}}$ of each spectrum 
    is known, the according blackbody spectrum was divided out.
    \item \textit{Percentile Normalization}: The spectra were normalized by dividing
    them by their 99th percentile.
    \item \textit{Continuum Normalization}: The spectra were further normalized by 
    dividing them by a \nth{5} order polynomial continuum.
\end{enumerate}
Mathematically, we can express the preprocessing as follows:
\begin{gather}
    \mathsf{\bar{S}} = \dfrac{\mathsf{S}}{\mathsf{B}\mathsf{Q}_5P_{99}}
\end{gather}
where $\mathsf{\bar{S}}$ is the preprocessed spectrum, $\mathsf{S}$ is the 
original spectrum,  $\mathsf{B}$ is the blackbody spectrum, $\mathsf{Q}_n$ 
is the $n^\mathrm{th}$ order polynomial continuum fit, and $P_n$ is the 
$n^\mathrm{th}$ percentile function. Arithmetic operations between arrays 
are assumed to be elementwise in all notation from here on out.

\subsection{Line Identification with \texttt{blase}}
The next step was to convert the PHOENIX subset into an interpretable format.
We wanted to represent the spectra as a list of spectral lines rather than a
list of fluxes. This was done using \texttt{blase}, which detects spectral 
lines as Voigt profiles and tunes the profiles to mimic the original 
PHOENIX spectrum with back propagation. Four parameters were optimized: the 
line center $\mu$, the log-amplitude $\ln(a)$, the Gaussian width $\sigma$, and the 
Lorentzian width $\gamma$. The optimization used the Adam optimizer with a
learning rate of 0.05 over 100 epochs. In addition, we limited two custom 
parameters: wing cut to 6000 and prominence to 0.005. Wing cut is a parameter
that determines the extent of the Voigt profile to evaluate, saving
computational resources by not evaluating small numbers. Prominence sets a
lower limit for the amplitude of detected lines, which also saves resources 
by disregarding shallow lines. In short, larger values for wing cut and 
smaller values for prominence both increase the accuracy of using
\texttt{blase}'s line detection and optimization at the expense of an
increased computational cost. In addition, it should be mentioned that
\texttt{blase} uses the pseudo-Voigt approximation, which saves on 
computational cost while remaining accurate to about 1\%. It is a 
weighted average of a Gaussian and Lorentzian as opposed to a convolution.
\texttt{blase}'s pseudo-Voight profile implementation uses the following:
\begin{subequations}
\begin{gather}
    \tilde{\mathsf{V}}_\mu(\lambda) = a\left[\eta \mathbf{L}(\lambda - \mu'; f) + (1 - \eta)\mathbf{G}(\lambda - \mu'; f)\right]\\
    \eta = \sum_{n=1}^3 \mathbf{u}_n\left(\dfrac{2\gamma}{f}\right)^n\\
    f = 32\sum_{n=0}^5 \mathbf{v}_n \left(\sqrt{2\ln(2)}\sigma\right)^{5-n}(\gamma)^n\\
    \nonumber \mathbf{u} = \mqty[1.36603 \\ -0.47719 \\ 0.11116] \quad \mathbf{v} = \mqty[1 \\ 2.69269 \\ 2.42843 \\ 4.47163 \\ 0.07842 \\ 1]
\end{gather}
\end{subequations}
where $\mathbf{L}$ and $\mathbf{G}$ are abbreviations for Lorentzian and 
Gaussian profiles, respectively. Notice that we use $\mu'$ instead of $\mu$ 
in the formula. This is because \texttt{blase} allows the line center to shift 
slightly during optimization, and it is this shifted center which is used in 
computation. The individual Voigt profiles are still indexed by $\mu$ for 
cross-model line identification, explained in the next section. Once optimization 
was complete, the list of identified lines, present in the state dictionary of 
\texttt{blase}'s model, was saved to a \texttt{.pt} file for each PHOENIX subset 
grid point. The total disk space these files took up is 465 MB.

@@

\subsection{Inference Algorithm}
We elected to use Bayesian optimization as the inference algorithm, specifically 
the \texttt{gp\_minimize} function from the \texttt{scikit-optimize} library. 
This algorithm uses a Gaussian Process to model the objective function, which in 
this case was the RMS (Root-Mean-Square) loss between the interpolated spectrum 
$\mathsf{M}$ and the true spectrum $\mathsf{D}$, defined as:
\begin{gather}
    \mathcal{L} = \ev{(\mathsf{M} - \mathsf{D})^{\circ 2}}^{1/2}
\end{gather}
The optimizer was configured to first run 30 random 
evaluations to seed the surrogate model, then run 20 more evaluations now guided 
by the surrogate model. This totals to 50 evaluations, which was deemed 
sufficient for this study. One inference run took us an average of 15 minutes to 
complete.

@@

    \item \textit{Low-Order Interpolation}: The current design of the 
    algorithm uses linear interpolation to construct manifolds, however this
    could be made more sophisticated with more complex interpolation schemes.
    However, the drawback of this is that higher orders of interpolation
    become increasingly expensive both in terms of disk utilization, and 
    in terms of the computational cost to generate and evaluate them. Future
    studies should sidestep this issue by using more generalizable methods
    that are much more efficient.

@@

advent of physics-informed machine learning \citep{test} \citep{test2}.

@@

\software{Text}

\begin{thebibliography}{dummy}
    \bibitem[Test 2024()]{test} Woop
    \bibitem[Test2 2024()]{test2} Woop woop
\end{thebibliography}

>>>>

\documentclass[twocolumn]{aastex631}
\usepackage{graphicx}
\usepackage{physics}
\usepackage[super]{nth}
\usepackage{enumitem}

\graphicspath{{./images/}}
\makeatletter
\renewcommand\hyper@natlinkbreak[2]{#1}
\makeatother

@@

\begin{abstract}
    The advent of machine learning methods has revolutionized 
    numerous scientific methods, among which is spectroscopic 
    inference. Studies can and have shown that machine learning 
    methods prove effective in inferring what we call
    fundamental stellar parameters. These include $T_{\mathrm{eff}}$,
    $\log(g)$, and [Fe/H] for the purposes of this study but can 
    also encompass other parameters such as [$\alpha$/Fe], C/O, and
    $f_{sed}$. While current methods tend to treat spectra purely
    from a data perspective, we propose to use a hybrid machine 
    learning (HML) method that uses the semi-empirical approach 
    of treating spectra as a set of spectral lines put forth by 
    \texttt{blase} \citep{blase}, combined with traditional linear interpolation 
    as well as Bayesian optimization. We interpolate the four 
    line properties ($\mu'$, $\ln(a)$, $\sigma$, $\gamma$) of 
    128,723 unique spectral lines across a subset of the PHOENIX
    synthetic model grid \citep{PHOENIX} to create the PHOENIX generator, allowing
    us to reconstruct spectra at any continuously valued point 
    lying within the domain of the PHOENIX subset, and even reconstruct 
    batches of spectra in pseudo-parallel fashion, with the speedup 
    exceeding 20 times once more than 60 spectra are reconstructed. 
    We then combine the PHOENIX generator with Bayesian optimization to 
    efficiently infer fundamental stellar parameters. This study is 
    however a proof of concept, as we have not yet included various
    features that would be necessary for a full-fledged tool for 
    spectroscopic inference, however we show that semi-empirical 
    HML methods are a viable, albeit fledgling alternative to 
    more traditional approaches.
\end{abstract}

@@

Scientists want to extract as much information as possible from
a star's spectra to get back the properties that lie at the root of it all. 
However, this has proven to be no easy task. Spectroscopic surveys such as 
APOGEE develop their own in-house pipelines to extract stellar parameters
from spectra (ASPCAP in the case of APOGEE), and these pipelines are fairly
effective \citep{APOGEE, ASPCAP}. However, they all work on the core assumption that the data to be 
analyzed is a list of pixels. In addition, the pipelines that surveys 
develop are typically closed-source and usually limited to the scope of the 
survey itself, causing a sort of fragmentation. This motivates the 
development of a more universal, open-source tool that can do what spectral 
analysis pipelines do, but in an alternative way.

Multiple efforts have been made in this direction, treating spectra in 
different ways. The standard practice is to treat the wavelength and flux as 
simply two arrays and implement any from a myriad of different algorithms to 
inference fundamental stellar parameters. Others take a more mathematical 
approach by decomposing spectra into some eigenbasis and treating it as a 
regression problem, such as \texttt{starfish} \citep{starfish}. However, \texttt{blase} takes 
a different approach. It treats spectra not as a set of pixels or a set of 
eigenbasis coefficients but as a set of spectral lines, specifically Voigt 
profiles. There is no metric of this approach being better or worse than the 
others mentioned so far, but it is a different paradigm and is worth 
exploring as a semi-empirical tool. This study aims to take advantage of 
\texttt{blase}'s ability to change a spectrum's basis from flux versus 
wavelength into spectral line versus spectral line properties, and use that 
to develop a robust tool for stellar parameter inference.

@@

We chose the PHOENIX synthetic spectral model grid as the basis for this 
study, because it is fairly well-known. While we plan to extend this tool to 
leverage information across multiple model grids in the future, both for 
accuracy and to increase the scope to include substellar objects with grids
such as Sonora \citep{bobcat, cholla, diamondback, elfowl}, for now we limit our scope to a subset of the PHOENIX grid.

@@

\begin{table}[h!]
    \centering
    \begin{tabular}{lll}
        \hline
        \bf{Parameter} & \bf{Symbol} & \bf{Range}\\
        \hline
        Alpha Element Abundance & $\alpha$ & [0] dex\\
        Iron Abundance & [Fe/H] & [-0.5, 0] dex\\
        Effective Temperature & $T_{\mathrm{eff}}$ & [2300, 12000] K\\
        Surface Gravity & $\log(g)$ & [2, 6]\\
        Wavelength & $\lambda$ & [8038, 12849] \AA\\
        \hline
    \end{tabular}
    \caption{The subset of the PHOENIX grid used in this study. These limits 
    were imposed to reduce the computational cost of the algorithms and to 
    ensure a rectilinear parameter space in order to work with \texttt{scipy}'s
    \texttt{RegularGridInterpolator} \citep{scipy}. The wavelength limits in particular 
    roughly line up with that of the Habitable Zone Planet Finder (HPF) 
    spectrograph \citep{HPF}. This subset is comprised of 1314 individual spectra.}
\end{table}

\subsection{Preprocessing with \texttt{gollum}}
First, the PHOENIX subset was accessed directly from the PHOENIX website
using \texttt{gollum}'s download option \citep{gollum}. The spectra were then put through a 
three-step preprocessing pipeline.
\begin{enumerate}
    \item \textit{Blackbody Division}: Since the $T_{\mathrm{eff}}$ of each spectrum 
    is known, the according blackbody spectrum was divided out.
    \item \textit{Percentile Normalization}: The spectra were normalized by dividing
    them by their 99th percentile.
    \item \textit{Continuum Normalization}: The spectra were further normalized by 
    dividing them by a \nth{5} order polynomial continuum.
\end{enumerate}
Mathematically, we can express the preprocessing as follows:
\begin{gather}
    \mathsf{\bar{S}} = \dfrac{\mathsf{S}}{\mathsf{B}\mathsf{Q}_5P_{99}}
\end{gather}
where $\mathsf{\bar{S}}$ is the preprocessed spectrum, $\mathsf{S}$ is the 
original spectrum,  $\mathsf{B}$ is the blackbody spectrum, $\mathsf{Q}_n$ 
is the $n^\mathrm{th}$ order polynomial continuum fit, and $P_n$ is the 
$n^\mathrm{th}$ percentile function. Arithmetic operations between arrays 
are assumed to be elementwise in all notation from here on out.

\subsection{Line Identification with \texttt{blase}}
The next step was to convert the PHOENIX subset into an interpretable format.
We wanted to represent the spectra as a list of spectral lines rather than a
list of fluxes. This was done using \texttt{blase}, which detects spectral 
lines as Voigt profiles and tunes the profiles to mimic the original 
PHOENIX spectrum with back propagation. Four parameters were optimized: the 
line center $\mu$, the log-amplitude $\ln(a)$, the Gaussian width $\sigma$, and the 
Lorentzian width $\gamma$. The optimization used the Adam optimizer with a
learning rate of 0.05 over 100 epochs \citep{adam}. In addition, we limited two custom 
parameters: wing cut to 6000 and prominence to 0.005. Wing cut is a parameter
that determines the extent of the Voigt profile to evaluate, saving
computational resources by not evaluating small numbers. Prominence sets a
lower limit for the amplitude of detected lines, which also saves resources 
by disregarding shallow lines. In short, larger values for wing cut and 
smaller values for prominence both increase the accuracy of using
\texttt{blase}'s line detection and optimization at the expense of an
increased computational cost. In addition, it should be mentioned that
\texttt{blase} uses the pseudo-Voigt approximation, which saves on 
computational cost while remaining accurate to about 1\% \citep{pseudovoigt1, pseudovoigt2}. It is a 
weighted average of a Gaussian and Lorentzian as opposed to a convolution.
\texttt{blase}'s pseudo-Voight profile implementation uses the following:
\begin{subequations}
\begin{gather}
    \tilde{\mathsf{V}}_\mu(\lambda) = a\left[\eta \mathbf{L}(\lambda - \mu'; f) + (1 - \eta)\mathbf{G}(\lambda - \mu'; f)\right]\\
    \eta = \sum_{n=1}^3 \mathbf{u}_n\left(\dfrac{2\gamma}{f}\right)^n\\
    f = 32\sum_{n=0}^5 \mathbf{v}_n \left(\sqrt{2\ln(2)}\sigma\right)^{5-n}(\gamma)^n\\
    \nonumber \mathbf{u} = \mqty[1.36603 \\ -0.47719 \\ 0.11116] \quad \mathbf{v} = \mqty[1 \\ 2.69269 \\ 2.42843 \\ 4.47163 \\ 0.07842 \\ 1]
\end{gather}
\end{subequations}
where $\mathbf{L}$ and $\mathbf{G}$ are abbreviations for Lorentzian and 
Gaussian profiles, respectively. Notice that we use $\mu'$ instead of $\mu$ 
in the formula. This is because \texttt{blase} allows the line center to shift 
slightly during optimization, and it is this shifted center which is used in 
computation. The individual Voigt profiles are still indexed by $\mu$ for 
cross-model line identification, explained in the next section. Once optimization 
was complete, the list of identified lines, present in the state dictionary of 
\texttt{blase}'s model, was saved to a \texttt{.pt} file for each PHOENIX subset 
grid point \citep{pytorch}. The total disk space these files took up is 465 MB.

@@

\subsection{Inference Algorithm}
We elected to use Bayesian optimization as the inference algorithm, specifically 
the \texttt{gp\_minimize} function from the \texttt{scikit-optimize} library \citep{skopt}. 
This algorithm uses a Gaussian Process to model the objective function, which in 
this case was the RMS (Root-Mean-Square) loss between the interpolated spectrum 
$\mathsf{M}$ and the true spectrum $\mathsf{D}$, defined as:
\begin{gather}
    \mathcal{L} = \ev{(\mathsf{M} - \mathsf{D})^{\circ 2}}^{1/2}
\end{gather}
The optimizer was configured to first run 30 random 
evaluations to seed the surrogate model, then run 20 more evaluations now guided 
by the surrogate model. This totals to 50 evaluations, which was deemed 
sufficient for this study. One inference run took us an average of 15 minutes to 
complete.

@@

@@

advent of physics-informed machine learning.

@@

\software{
    \texttt{altair} \citep{altair1, altair2},
    \texttt{astropy} \citep{astropy1, astropy2, astropy3},
    blas\'e/\texttt{blase} \citep{blase},
    CUDA \citep{cuda},
    \texttt{gollum} \citep{gollum},
    \texttt{matplotlib} \citep{mpl},
    \texttt{numpy} \citep{numpy},
    \texttt{pandas} \citep{pandas1, pandas2},
    Python \citep{python},
    PyTorch/\texttt{torch} \citep{pytorch},
    scikit-optimize/\texttt{skopt} \citep{skopt},
    \texttt{scipy} \citep{scipy},
    \texttt{tqdm} \citep{tqdm},
    \texttt{vegafusion} \citep{vegafusion},
}

\bibliography{paper}
